---
output:
  md_document:
    variant: gfm

---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = F}
knitr::opts_chunk$set(echo = TRUE)
options(tidyverse.quiet = TRUE)
library(volatility)
```
## Introduction
Volatility refers to the degree of instability (or change over time) in the microbiome. High volatility, ie an unstable microbiome, has been associated with an exaggerated stress response and conditions like IBS. 

This library provides a basic framework to calculate volatility for timepoint microbiome data. 

If you use this software, please cite our work.

Thomaz F.S Bastiaanssen, Anand Gururajan, Marcel van de Wouw, Gerard M Moloney, Nathaniel L Ritz, Caitriona M Long-Smith, Niamh C Wiley, Amy B Murphy, Joshua M Lyte, Fiona Fouhy, Catherine Stanton, Marcus J Claesson, Timothy G Dinan, John F Cryan,
**Volatility as a Concept to Understand the Impact of Stress on the Microbiome**
Psychoneuroendocrinology
https://doi.org/10.1016/j.psyneuen.2020.105047

## Setup

OK, now let's get started.
We'll load a complementary training dataset using `data(volatility_data)`. This loads a curated snippet from the dataset described in more detail here: https://doi.org/10.1016/j.psyneuen.2020.105047

The method presented here is differs only marginally from the one used there. 


```{r, eval=TRUE}
#install and load volatility library
#devtools::install_github("thomazbastiaanssen/volatility")
library(volatility)

#load tidyverse to wrangle and plot results.
library(tidyverse)

#load example data + metadata from the volatility study.
data(volatility_data)

```

## Input data

The main `volatility` function does all the heavy lifting here. It expects two objects:

* `counts`, a microbiome feature count table, with columns as samples and rows and features. 
    + The `vola_genus_table` object is an example of an appropriately formatted count table.
```{r, eval=TRUE}
vola_genus_table[4:10,1:2]
```

* `metadata`, a vector in the same order as the count table, denoting which samples are from the same source. 
    + The column `ID` in `vola_metadata` is appropriate for this. 
    
```{r, eval=TRUE}
head(vola_metadata)

```

## Basic use

```{r, eval=TRUE}
vola_out <- volatility(counts = vola_genus_table, metadata = vola_metadata$ID)

head(vola_out)
```

The output of the main `volatility` function is a data.frame with two columns. `ID` corresponds to the pairs of samples passed on in the `metadata` argument, whereas `volatility` shows the measured volatility between those samples in terms of Aitchison distance (Euclidean distance of CLR-transformed counts). 

## Plot the results


```{r plot_volatility, eval=TRUE, fig.width = 5, fig.height = 5}
vola_out %>%
  #Merge the volatilty output with the rest of the initial metadata using the shared "ID" column
  left_join(vola_metadata[vola_metadata$timepoint == "Pre",], "ID") %>%
  
  #Pipe into ggplot
  ggplot(aes(x = treatment, y = volatility, fill = treatment)) +
  
  #Define geoms, boxplots overlayed with data points in this case
  geom_boxplot(alpha = 1/2)+
  geom_point(shape = 21) +
  
  #Split the plot by cohort
  facet_wrap(~cohort) +
  
  #Tweak appearance 
  scale_fill_manual(values = c("Control" = "#3690c0", "Stress"  = "#cb181d")) +
  theme_bw() +
  xlab("") +
  ylab("Volatility (Aitchison distance)")
```

